\plaintitle

### 2016-11-20 Paper/Rebuttal.txt

We thank the reviewers for their thoughtful feedback. We are excited to present this work at CHI, and believe that addressing the issues raised in the reviews will make the work a stronger submission with greater utility to the research audience.

If accepted, in addition to the more minor fixes for issues presented by the reviewers, we intend to revise three key areas, as per the guidance of the meta review:

1) Clarification of the statistics used in the paper:

We stand by our decision to use interquartile means and bootstrapped confidence intervals. In addition to [8], these method for showing effect sizes have also been used in more recent graphical perception papers, such as [29]. We will expand our justification for these techniques in the paper, and  clarify our methodology.

IQMs: 
[8] (in section 4.4) uses “the means of the midmeans [IQM] of the log absolute errors” in their charts, including those in the caterpillar plot in [8] Fig. 16 (which R3 is correct in citing as a model for the figures in this paper). R3 correctly notes that the IQM has some disadvantages. In particular, the distribution of such values tends to violate the assumptions of normality of sampling error, which is key for tests such as t-tests and ANOVAs. In the paper, we use the IQM only for plots and for reporting effect sizes; for ANOVAs and other inferential tests, we test the standard mean. We will clarify this distinction in the paper.

Bootstrapped CIs:
Bootstrapped c.i.s appear frequently in graphical perception work, e.g. in [8,20,29]. [8] (also in section 4.4) justifies the use of bootstrapping as a robust and non-parametric way of creating c.i.s.

While we reiterate that the p-values reported in the paper were generated from standard means rather than the IQM, a more troubling issue with IQMs is that they reduce the variance of the variables of interest, and so can artificially inflate effect sizes, or create effects where none exist with other measures of central tendency. To set both meta, R3, and our own minds at ease, we generated CIs using a more standard approach, a two-tailed 95% t-confidence interval about the mean. While these CIs result in smaller effect sizes (due to higher variance), they do not alter any of the main results of the paper.

e.g., Figure 6 would be:
sigma, mean(absolute error) +/- 95%t-ci
0.05,  0.04 +/- (0.004)
0.10,  0.10 +/- (0.005)
0.15,  0.15 +/- (0.006)
0.20,  0.23 +/- (0.01)

Figure 8 would be:
graph type, mean(error) +/- 95%t-ci
area,  -0.01 +/- (0.006)
line,  0.01 +/- (0.005)
scatterplot, 0.005 +/- (0.003)

Figure 11 would be:
error type, mean(absolute error) +/- 95%t-ci
w/r/t OLS fit, 0.4 +/- 0.009
w/r/t robust fit, 0.15 +/- 0.007
no outliers, 0.1 +/- 0.006

and so on. While the data are noisier using this procedure, the trends (monotonic increase in error as residuals get larger, “within-the-bar bias” underestimation in area charts but not other chart types, participants are closer to the robust fit than the OLS fit, etc.) remain.

2) Additional detail about experimental methods:

We will include or clarify study details (number of exclusions, re-structuring of the methods section to make the participant information clearer, etc.) asked for by the reviewers.

Additionally, while we have already included our study data in our supplemental materials, on acceptance, we would make public our (currently private) repository containing not only data tables and analyses, but also experimental stimuli and instruments. In the main body of the paper, we will include figures with more example stimuli. We can include additional figures (such as violin plots of per-subject IQMs, or other non-parametric visualizations of effect size) in the online supplement as well.

R4 and R5 mention, as an alternative study design, un-constrained input of trend lines, rather than our study design (which is more concerned with fitting a given model, linear or otherwise, than the freeform estimation of a fit, which may or may not correspond to any of the categories of provided models). For the targeted research questions in our study, we believe our more parsimonious study design is justified. However, our group is actively researching the design and implementation of experimentally tractable freeform solicitation techniques (for instance, the “draw your guess” NYT example mentioned by R5). We will clarify the distinction between these two types of visual regression, and cite work in this area that has been published since our initial submission.

3) Better situation of this work within the CHI community, and better communication of its utility:

While many graphical perception works have had a good reception at CHI (Fuchs et al “Evaluation of Alternative Glyph Designs for Time Series Data in a Small Multiple Setting” and [10] were presented at CHI 2013 and 2012, respectively), we believe that we can improve the perceived fit of this work at CHI through the inclusion of an explicit implications for design subsection, as part of the conclusion (as per R5). We believe the strength of our results offers guidance to designers in information visualization: in particular, it seems to suggest that area charts (which are a common encoding for time series data) have unanticipated downsides when attempting to convey temporal trends. Similarly, it suggest that standard trend lines (which are OLS lines by default in programs like Excel and standard statistical toolkits) may not appear to be credible if there are many outliers in the data; in such cases, it appears that viewers can be trusted to perform the robust fits visually.

This issue of trust in the viewer (or the stats!) to be “well behaved” when it comes to estimating statistical properties, or making use of visualizations, is central to how we developed our research questions for this paper, and we will expand upon it in our introduction (particular our Anscombe quartet example, as per R5), and in scoping our claimed contributions.

We hope that this rebuttal addresses the main concerns brought up by the reviewers, and would like to thank them again for their helpful and thorough comments. As the revisions we propose above are largely to the argumentation and presentation of the paper, we are confident that we can carry them out within the CHI revision cycle, and apply them to the final paper.

### 2016-11-20 Paper/Rebuttal.txt

We thank the reviewers for their thoughtful feedback. We hope to present this work at CHI, and believe that addressing the issues raised in the reviews will make the work a stronger submission with greater utility. If accepted (in addition to other, more minor changes suggested by the reviewers), as per the meta-review we will revise three key areas:

1) Clarification of the statistical methods:

We stand by our decision to present midmeans and bootstrapped confidence intervals. In addition to [8], these methods for showing effect sizes have also been used in more recent graphical perception papers [17, 29]. We will expand our justification for these techniques in the paper, and clarify our methodology.

Bootstrapped CIs:
Bootstrapped CIs appear frequently in graphical perception work [8, 17, 29]. Prior work ([8], section 4.4) justifies the use of bootstrapping as a robust and non-parametric way of creating CIs.

Midmeans (aka Interquartile Means, or IQMs):
Prior work [8, 17] indeed uses "the means of the midmeans of the log absolute errors" in charts, including Fig. 16 in [8] (which R3 correctly cites as a model for figures in our paper). R3 notes some disadvantages of IQMs: the distribution of such values tends to violate the assumptions of normality of sampling error, which is key for t-tests and ANOVAs. In the paper, we use IQMs only for plots and for reporting effect sizes; for ANOVAs and other inferential tests, we use standard means. We will clarify this distinction in the paper.

While we reiterate that the p-values reported in the paper come from standard means (not IQMs), it is true that IQMs reduce the variance of the variables of interest, and so may artificially inflate effect sizes or create effects unseen with other measures of central tendency. In response to this concern, we generated CIs using a more standard approach: a two-tailed 95% t-confidence interval of the standard mean. While these CIs result in smaller effect sizes (due to higher variance), they do not alter any of the main results of the paper.

Figure 6 becomes:
sigma, mean(absolute error) +/- 95%t-ci
0.05,  0.04 +/- (0.004)
0.10,  0.10 +/- (0.005)
0.15,  0.15 +/- (0.006)
0.20,  0.23 +/- (0.01)

Figure 8 becomes:
graph type, mean(error) +/- 95%t-ci
area,  -0.01 +/- (0.006)
line,  0.01 +/- (0.005)
scatterplot, 0.005 +/- (0.003)

Figure 11 becomes:
error type, mean(absolute error) +/- 95%t-ci
w/r/t OLS fit, 0.4 +/- 0.009
w/r/t robust fit, 0.15 +/- 0.007
no outliers, 0.1 +/- 0.006

...and so on. While estimates are noisier under this procedure, the results remain unchanged: monotonic increase in error as residuals get larger, “within-the-bar bias” underestimation in area charts but not other chart types, participants are closer to the robust fit than the OLS fit, etc.

@@

2) Additional detail about experimental methods:

We will include or clarify study details asked for by reviewers, including the number of exclusions, and re-structuring of the methods section to make the participant information clearer.

Additionally, though we did submit our study data as supplemental material, on acceptance, we would make public our full (currently private) repository containing not only data tables and analyses, but also experimental stimuli and instruments. In the main body of the paper, we will add figures with more example stimuli. We can also include additional figures (such as violin plots of per-subject IQMs, or other non-parametric visualizations of effect size) in the online supplement.

As an alternative to our study design, both R4 and R5 mention un-constrained input of trend lines (which we had also initially considered). Our design is more concerned with fitting a given model (linear or otherwise) than a freeform "sketch" of a fit, which may or may not correspond to any of the categories of provided models -- and may also suffer from additional noise such as drawing "jitter" or "satisficing" by subjects due to the higher cost of revising drawings. For the targeted research questions in our study, we believe our more parsimonious study design is justified. However, we know of others actively researching the design and implementation of experimentally tractable freeform solicitation techniques (for instance, the “draw your guess” NYT example mentioned by R5). We will clarify the distinction between these two types of visual regression, citing work in this area that has been published since our initial submission.

3) Better situate the work in the CHI community, and better communicate its utility:

While many graphical perception works have had a welcome reception at CHI (e.g., [10, 17]), we believe that we can improve the perceived fit of this work at CHI (as per R5) through the inclusion of an explicit implications for design subsection of the conclusion. We believe our results offer guidance to designers in information visualization. In particular, they suggest that area charts (which are a common encoding for time series data) have unanticipated downsides when attempting to convey temporal trends. Similarly, they suggest that standard trend lines (which are OLS lines by default in programs like Excel and other common analysis tools) may not appear to be credible if there are many outliers in the data; in such cases, it appears that viewers can be trusted to perform the robust fits visually.

This issue of trust in the viewer (or the stats!) to be "well behaved" when it comes to estimating statistical properties, or making use of visualizations, is central to how we developed our research questions for this paper, and we will expand upon it in our introduction (particular our Anscombe quartet example, as per R5), and in scoping our claimed contributions.

We hope our rebuttal addresses the main reviewer concerns, and would like to thank the reviewers again for their helpful and thorough comments. As the revisions we propose above largely concern argumentation and presentation issues, we are confident that we can carry them out within the CHI revision cycle.

### 2016-11-20 Paper/Rebuttal.txt

We thank the reviewers for their thoughtful feedback. We hope to present this work at CHI, and believe that addressing the issues raised in the reviews will make the work a stronger submission. If accepted (in addition to other, more minor changes suggested by the reviewers), as per the meta-review we will revise three key areas:

@@

We stand by our decision to present midmeans and bootstrapped confidence intervals. In addition to [8], these methods for showing effect sizes have also been used in more recent graphical perception papers [17, 29]. We will expand our justification for these techniques in the paper, and clarify our methodology.

Bootstrapped CIs:
Bootstrapped CIs appear frequently in graphical perception work [8, 17, 29]. Prior work ([8], section 4.4) justifies the use of bootstrapping as a robust and non-parametric way of creating CIs.

Midmeans (aka Interquartile Means, or IQMs):
Prior work [8, 17] indeed uses "the means of the midmeans of the log absolute errors" in charts, including Fig. 16 in [8] (which R3 correctly cites as a model for figures in our paper). R3 notes some disadvantages of IQMs: the distribution of such values tends to violate the assumptions of normality of sampling error, which is key for t-tests and ANOVAs. In the paper, we use IQMs only for plots and for reporting effect sizes; for ANOVAs and other inferential tests, we use standard means. We will clarify this distinction in the paper.

While we reiterate that the p-values reported in the paper come from standard means (not IQMs), IQMs have tend to have lower variance than standard means, and so may inflate effect sizes compared with other measures of central tendency. In response to this concern, we generated two-tailed 95% t-confidence intervals of the standard mean. These new CIs do not alter the main results of the paper.

Fig 6 becomes:
sigma, mean(absolute error)+/-95%t-ci
0.05,  0.04 +/- (0.004)
0.10,  0.10 +/- (0.005)
0.15,  0.15 +/- (0.006)
0.20,  0.23 +/- (0.01)

Fig 8 becomes:
graph type, mean(error)+/-95%t-ci
area,  -0.01 +/- (0.006)
line,  0.01 +/- (0.005)
scatterplot, 0.005 +/- (0.003)

Fig 11 becomes:
error type, mean(absolute error)+/-95%t-ci
w/r/t OLS fit, 0.4 +/- 0.009
w/r/t robust fit, 0.15 +/- 0.007
no outliers, 0.1 +/- 0.006

...and so on. While effect sizes are smaller, the results remain unchanged: increase in error as residuals get larger, “within-the-bar bias” underestimation in area charts, estimations closer to the robust fit than the OLS fit, etc.

@@

2) Additional detail about experimental methods:

We will include or clarify study details asked for by reviewers, including the number of exclusions, and re-structuring of the methods section to make the participant information clearer. Though we did submit our study data as a supplement, on acceptance, we would make public our full repository containing data tables, analyses, experimental stimuli, and instruments.

As an alternative study design, both R4 and R5 mention un-constrained input of trend lines (which we had also initially considered). Our design is more concerned with fitting a given model (linear or otherwise) than a freeform "sketch" of a fit, which may or may not correspond to any of the categories of provided models -- and may also suffer from additional noise such as drawing "jitter" or "satisficing" by subjects due to the higher cost of revising drawings. For the targeted research questions in our study, we believe our more parsimonious study design is justified. However, we know of others actively researching the design of experimentally tractable freeform solicitation techniques (e.g., the “draw your guess” NYT example mentioned by R5). We will clarify the distinction between these types of task, citing work in this area that has been published since our submission.

3) Better situate the work in the CHI community:

While many graphical perception works have had a welcome reception at CHI (e.g., [10, 17]), we believe that we can improve the perceived fit of this work at CHI (as per R5) through the inclusion of an implications for design subsection of the conclusion. We believe our results offer guidance to designers in visualization. In particular, they suggest that area charts (a common encoding for time series data) have unanticipated downsides when attempting to convey temporal trends. Similarly, they suggest that standard trend lines (which are OLS lines by default in programs like Excel and other common analysis tools) may not appear credible if there are many outliers in the data; in such cases, viewers could be trusted to perform the robust fits visually.

### 2017-01-02 Paper/proceedings.tex
Draft final submission.

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
% \toappear{Permission to make digital or hard copies of all or part of this work for personal or classroom use is      granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
% DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

@@

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Regression by Eye: Estimating Trends\\in Bivariate Visualizations}
\def\plainauthor{Removed for Review}
\def\emptyauthor{}
\def\plainkeywords{Information Visualization, Graphical Perception, Regression}
\def\plaingeneralterms{Visualization}

@@

\numberofauthors{2}
\author{%
%  \alignauthor{Michael Correll\\
%    \affaddr{University of Washington}\\
%    \email{mcorrell@cs.washington.edu}}\\
%  \alignauthor{Jeffrey Heer\\
%    \affaddr{University of Washington}\\
%    \email{jheer@cs.washington.edu}}\\
}

@@

\begin{abstract}
Observing trends and predicting future values are common tasks for viewers of bivariate data visualizations. As many charts do not explicitly include trend lines or related statistical summaries, viewers often visually estimate trends directly from a plot. How reliable are the inferences viewers draw when performing such \emph{regression by eye}? Do particular visualization designs or data features bias trend perception? We present a series of crowdsourced experiments that assess the accuracy of trends estimated using regression by eye across a variety of bivariate visualizations, and examine potential sources of bias in these estimations. We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within the bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models.
% In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge. These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
% JEFF: These last two lines largely re-state the preceding sentence. Maybe add a new line to the abstract that summarizes implications for visualization designers or tool makers.
\end{abstract}

@@

In order to assess the ability of visualization viewers to estimate trends, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk. We designed these experiments to establish a performance baseline for regression by eye, and to examine potential sources of bias. Crowdsourced graphical perception experiments have been found to produce results that are largely in keeping with prior, lab-based work~\cite{heer2010crowdsourcing, talbot2014four}. In this section, we describe experimental design aspects shared across all experiments.

We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%. For validation, additional trials, without ambiguities such as outliers or design differences, contained the actual (OLS) trend line, which subjects then needed to simply match.  We excluded the results of participants with average error greater than 0.2 (20 times the step size of the slider) on these validation questions. Across the three reported experiments and pilots, we performed 7 such exclusions. We recruited additional participants to replace these excluded subjects. Based on timings from internal piloting, we paid each participant \$2 for their participation, for a target rate of \$8/hour.

@@

We investigated three types of common bivariate visualizations: scatter plots, line graphs, and area charts (see Fig. \ref{fig:types}). In each experiment, we presented participants with a visualization and asked them to interactively adjust trend lines to best fit the data.  Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}. Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed). We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. We avoided more expressive designs (such as freeform drawing) in order to constrain the model space and to reduce the impact of factors such as noise from motor movements or heterogeneous input devices.

@@

@@

In this experiment, the primary measure of performance was the interquartile mean (or midmean) of absolute error. The interquartile mean discards points in the first and fourth quartiles before averaging, trimming the tails of distributions. Cleveland \& McGill~\cite{cleveland1984graphical} used the interquartile mean to provide a more robust measure of central tendency for responses from graphical perception studies, where participant error can create long tailed distributions of error. This is especially the case for crowdsourced studies, where correcting for data quality issues is perceived as more difficult than in laboratory settings~\cite{heer2010crowdsourcing}.

@@

%The average absolute error of slope estimation across all conditions had an interquartile mean of $0.06$ for slope values ranging from $\{-1,1\}$, corresponding to an error rate of $17\%$ of the actual slopes. 
Prior work in graphical perception often measures performance as the absolute log error: e.g., Cleveland \& McGill~\cite{cleveland1984graphical} calculate their performance metric as $\log_2 (|\text{error}| + \frac{1}{8}$). As a point of comparison, the interquartile mean of the absolute log error across all conditions in this experiment was $2.4$. Differences in methodology and measurement discourage statistical inferences comparing this value with those in similar experiments. Nevertheless, error rates for regression by eye are at least in the same magnitude as those observed in other graphical perception tasks in both lab and crowdsourced studies. For instance, Heer \& Bostock~\cite{heer2010crowdsourcing} report an absolute log error of $2.5$ for proportional judgments in treemaps, and Cleveland \& McGill~\cite{cleveland1984graphical} report an absolute log error of $2.4$ for proportional judgments in stacked bars (although compare to a log error of estimation of relative lengths of lines of $1.1$). This similarity suggests that, despite requiring estimation of aggregate statistical information, regression by eye results in judgments that are accurate enough for many practical purposes, comparable with the expected accuracy in comparing individual values in visualizations.

@@

``Within the bar'' bias is a known perceptual bias involving bar charts, in which points contained in the visual area of the glyph of the bar are deemed likelier than points outside of the glyph. Newman \& Scholl~\cite{newman2012bar} encountered this bias for a sampling task: ``how likely is this point to have been drawn from the distribution represented by this bar?'' Correll \& Gleicher~\cite{correll2014error} likewise encountered this bias for inferential tasks: ``how likely is the population mean to take a particular value, given the sample represented by this bar?''

We hypothesized that this bias would likewise occur in regression by eye when using visually asymmetric visualizations such as area charts. The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} under-estimation in trends; decreasing the slope would cause under-estimation at the beginning of the plot but not the end, and vice versa. We therefore designed this experiment to elicit estimates of the y-intercepts of trends. A within-the-bar bias would then appear as systematic under-estimation of intercept in area charts.

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends in area charts, as opposed to line graphs and scatter plots.
\end{enumerate}

@@

OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers~\cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values will continue) can impact how viewers categorize and utilize outliers in prediction tasks~\cite{campbell2009anchoring, ji2001culture}. In other words, regression by eye may not uniformly weight outliers, depending on their position in the plot.

@@

Our results fail to support our second hypothesis. \textbf{While participant estimates diverge from the robust line as the number of outliers increases, the OLS trend line diverges faster than participant estimates}. There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$, $F(1,4591)=110$, $p<0.001$). However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$, $F(1,4591)=2800$, $p<0.001$). Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line monotonically increased almost 7-fold, from $0.07$ with 0 outliers to $0.54$ with 15 outliers. These results indicate that, while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.  Figure \ref{fig:exp3outliers} illustrates this result.

@@

However, this general accuracy of trend estimation is not universal. Area charts are visually asymmetrical: the area below the line is filled in, and the area above it is not. We found that this asymmetry creates a within the bar bias: a systematic under-estimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area. Designers hoping to rely on regression by eye should avoid such asymmetries in their bivariate visualizations.

@@

@@

In this paper, we examine the ability of visualization viewers to perform \emph{regression by eye}: the visual estimation of trends in bivariate visualizations. We show that viewers without statistical training can reliably estimate both linear and non-linear trends in charts such as scatter plots and line graphs. However, area graphs are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations. Regression by eye is also less sensitive to outliers than standard least-squares regression. This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by regression methods, that can be heavily influenced by a few extreme values.

@@

Omitted for review.

>>>>

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

 \toappear{\scriptsize Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
 	{\emph{CHI 2017, May 6-11, 2017, Denver, CO, USA.} } \\
 	Copyright \copyright~2017 ACM ISBN 978-1-4503-4655-9/17/05\ ...\$15.00. \\
 	http://dx.doi.org/10.1145/3025453.3025922}
% DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

@@

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Regression by Eye: Estimating Trends\\in Bivariate Visualizations}
\def\plainauthor{Michael Correll and Jeffrey Heer}
\def\emptyauthor{}
\def\plainkeywords{Information Visualization, Graphical Perception, Regression}
\def\plaingeneralterms{Visualization}

@@

\numberofauthors{2}
\author{%
  \alignauthor{Michael Correll\\
    \affaddr{University of Washington}\\
    \email{mcorrell@cs.washington.edu}}\\
  \alignauthor{Jeffrey Heer\\
    \affaddr{University of Washington}\\
    \email{jheer@cs.washington.edu}}\\
}

@@

\begin{abstract}
Observing trends and predicting future values are common tasks for viewers of bivariate data visualizations. As many charts do not explicitly include trend lines or related statistical summaries, viewers often visually estimate trends directly from a plot. How reliable are the inferences viewers draw when performing such \emph{regression by eye}? Do particular visualization designs or data features bias trend perception? We present a series of crowdsourced experiments that assess the accuracy of trends estimated using regression by eye across a variety of bivariate visualizations, and examine potential sources of bias in these estimations. We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within-the-bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models.
% In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge. These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
% JEFF: These last two lines largely re-state the preceding sentence. Maybe add a new line to the abstract that summarizes implications for visualization designers or tool makers.
\end{abstract}

@@

In order to assess the ability of visualization viewers to estimate trends, we conducted a series of three crowdsourced experiments on Amazon's Mechanical Turk. We designed these experiments to establish a performance baseline for regression by eye, and to examine potential sources of bias. Crowdsourced graphical perception experiments have been found to produce results that are largely in keeping with prior, lab-based work~\cite{heer2010crowdsourcing, talbot2014four}. In this section, we describe experimental design aspects shared across all experiments. Data tables, stimuli, and experimental instruments are available online at \url{https://github.com/uwdata/trend-bias}.

In these experiments, we report effect size using the interquartile mean (or midmean) of absolute error. The interquartile mean discards points in the first and fourth quartiles before averaging, trimming the tails of distributions. Cleveland \& McGill~\cite{cleveland1984graphical} used the interquartile mean to provide a more robust measure of central tendency for responses from graphical perception studies, where participant error can create long tailed distributions of error. This is especially the case for crowdsourced studies, where correcting for data quality issues is perceived as more difficult than in laboratory settings~\cite{heer2010crowdsourcing}. However, the use of interquartile means can violate the assumptions involved in performing null hypothesis significance tests (as sample means can have non-normal distributions). Our statistical analyses were therefore performed using standard group means.

@@

We investigated three types of common bivariate visualizations: scatter plots, line graphs, and area charts (see Fig. \ref{fig:types}). In each experiment, we presented participants with a visualization and asked them to interactively adjust trend lines to best fit the data.  Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}. Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed). We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. While other studies investigate the use of more expressive designs (such as free-form drawing)~\cite{kim2017}, we sought to disambiguate the model selection task (e.g., the selection of a a linear or non-linear model) from the model fitting task (adjusting the parameters of the selected model). Our design also reduces impact of factors such as noise from motor movements or heterogeneous input devices.

@@

We limited the Mechanical Turk participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%. For validation, additional trials, without ambiguities such as outliers or design differences, contained the actual (OLS) trend line, which subjects then needed to simply match.  We excluded responses from participants with high average error (greater than 0.2) on these validation stimuli. Across the three reported experiments and pilots, we performed 7 such exclusions. We recruited additional participants to replace these excluded subjects. Based on timings from internal piloting, we paid each participant \$2 for their participation, for a target rate of \$8/hour.

@@

@@

Prior work in graphical perception often measures performance as the absolute log error: e.g., Cleveland \& McGill~\cite{cleveland1984graphical} calculate their performance metric as $\log_2 (|\text{error}| + \frac{1}{8}$). As a point of comparison, the interquartile mean of the absolute log error across all conditions in this experiment was $2.4$. Differences in methodology and measurement discourage statistical inferences comparing this value with those in similar experiments. Nevertheless, error rates for regression by eye are at least in the same magnitude as those observed in other graphical perception tasks in both lab and crowdsourced studies. For instance, Heer \& Bostock~\cite{heer2010crowdsourcing} report an absolute log error of $2.5$ for proportional judgments in treemaps, and Cleveland \& McGill~\cite{cleveland1984graphical} report an absolute log error of $2.4$ for proportional judgments in stacked bars (although compare to a log error of estimation of relative lengths of lines of $1.1$). This similarity suggests that, despite requiring estimation of aggregate statistical information, regression by eye results in judgments that are accurate enough for many practical purposes, comparable with the expected accuracy in comparing individual values in visualizations.

@@

``Within-the-bar'' bias is a known perceptual bias involving bar charts, in which points contained in the visual area of the glyph of the bar are deemed likelier than points outside of the glyph. Newman \& Scholl~\cite{newman2012bar} encountered this bias for a sampling task: ``how likely is this point to have been drawn from the distribution represented by this bar?'' Correll \& Gleicher~\cite{correll2014error} likewise encountered this bias for inferential tasks: ``how likely is the population mean to take a particular value, given the sample represented by this bar?''

We hypothesized that this bias would likewise occur in regression by eye when using visually asymmetric visualizations such as area charts: a ``within-the-area'' bias. The slope estimation task in the previous experiment would not capture this bias, as there is no method for participants to indicate a \emph{uniform} under-estimation in trends; decreasing the slope would cause under-estimation at the beginning of the plot but not the end, and vice versa. We therefore designed this experiment to elicit estimates of the y-intercepts of trends. A within-the-area bias would then appear as systematic under-estimation of intercept in area charts.

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within-the-area bias.} That is, participants would estimate lower values of y-intercepts of trends in area charts, as opposed to line graphs and scatter plots.
\end{enumerate}

@@

OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers~\cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that runs of high or low values will continue) can impact how viewers categorize and utilize outliers in prediction tasks~\cite{campbell2009anchoring, ji2001culture}. In other words, regression by eye may not uniformly weight outliers, depending on their position in the plot.

@@

Our results fail to support our second hypothesis. \textbf{While participant estimates diverge from the robust line as the number of outliers increases, the OLS trend line diverges faster than participant estimates}. There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$, $F(1,4591)=110$, $p<0.001$). However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$, $F(1,4591)=2800$, $p<0.001$). Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with $0$ outliers, monotonically increasing to $0.10$ with $15$ outliers), absolute error with respect to the OLS line monotonically increased almost 7-fold, from $0.07$ with $0$ outliers to $0.54$ with $15$ outliers. These results indicate that, while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.  Figure \ref{fig:exp3outliers} illustrates this result.

@@

However, this general accuracy of trend estimation is not universal. Area charts are visually asymmetrical: the area below the line is filled in, and the area above it is not. We found that this asymmetry creates a ``within-the-area'' bias: a systematic under-estimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area. Designers hoping to rely on regression by eye should avoid such asymmetries in their bivariate visualizations.

@@

\subsection{Implications for Design}

The estimation of trends, both for the prediction of values, and the imputation of missing values, is a common and important task in data analysis. We believe that our results suggest two main recommendations for designers of information visualizations that include time series or bivariate data:

\begin{enumerate}
	\item \emph{Designers do not need to annotate bivariate visualizations with trend lines in many cases}. Regression by eye, without explicit trend lines, is comparable in accuracy to other types of graphical perception tasks (such as comparison of value), even for a wide class of models. The presence of outliers is an exception to this general advice: human estimation of trends does not give much weight to outliers. For data where outliers are important, and expected to contribute to important trends, designers ought to visualize the outlier sensitive models directly.

	\item \emph{Designers should avoid area charts for data where the perception of trends is important.} The visual asymmetry caused by filling in the area under the line results in a corresponding asymmetry in judgment of trend: an underestimation in perceived trend. This within-the-area bias is undesirable in many cases, but can be countered by visual encodings (such as scatter plots and line charts) that have no such visual asymmetry.
\end{enumerate}

@@

In this paper, we examine the ability of visualization viewers to perform \emph{regression by eye}: the visual estimation of trends in bivariate visualizations. We show that viewers without statistical training can reliably estimate both linear and non-linear trends in charts such as scatter plots and line graphs. However, area graphs are subject to a ``within-the-area'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations. Regression by eye is also less sensitive to outliers than standard least-squares regression. This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by regression methods, that can be heavily influenced by a few extreme values.

@@

This work supported by a Moore Foundation Data-Driven Investigator Award.
